---
title: "3.2 Assignment"
author: "Scott Breitbach"
date: "9/13/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(ggplot2)
```

## 3.2 Assignment: Using Data to Improve a Marketing Promotion

For this week’s assignment we're going to use Dodgers Major League Baseball data from 2012. The data file you will be using is contained in the [dodgers.csv](http://content.bellevue.edu/cst/dsc/630/dodgers.csv) file. I would like you to determine what night would be the best to run a marketing promotion to increase attendance. It is up to you if you decide to recommend a specific date or if you  recommend a day of the week (e.g., Tuesdays) or month and day of the week (e.g., July Tuesdays). Use R and/or Python to accomplish this assignment. It is important to remember, there will be lots of ways to solve this problem. Explain your thought process and how you used various techniques to come up with your recommendation.

```{r load data}
# Load CSV file to Data Frame
df <- read.csv("dodgers.csv", header=TRUE)

# Get a preliminary look at the data
head(df)
summary(df)

# Create Histogram of Attendance since that is our key variable
hist(df$attend, xlab="Attendance", main="Histogram of Game Attendance")
```

The mean/median attendance values are pretty close, around 40-41 thousand, which will be our rough benchmark. I'll go through the variables in order to look for variations and to see if there is any indication of correlation with attendance that can be exploited to guide marketing promotions.  

First up: Month

```{r month}
# Assign month order so they chart correctly
df$month <- factor(df$month, levels=c("APR", "MAY", "JUN", "JUL", 
                                      "AUG", "SEP", "OCT"))
# Create boxplot
boxplot(attend ~ month, df, xlab="Month", ylab="Attendance", 
        main="Boxplots of Attendance by Month")
```

It appears the lowest months are May and October. If I had to choose one of those months to do a promotion, I would probably select May because it shows more variation, so a promotion may have a greater effect.

Up next: Day of the month. I wouldn't expect this to have any impact on attendance. If anything, one might see a slight uptick after the first and fifteenth of each month just because some people get paid on those days.

```{r day}
# First check for a correlation
print("Day/Attendance Correlation:")
cor(df$attend, df$day)

# Plot attendance by day of month
ggplot(df, aes(x=day, y=attend)) + geom_point() + geom_smooth(method="lm") + 
  labs(x="Day of Month", y="Attendance", title="Attendance by Day of Month") + 
  scale_x_continuous(n.breaks=20)
```

As expected, correlation is very low and there are no apparent patterns to the day of the month.

On to Day of the Week, where I expect to see some trends.

```{r day of week}
# Assign day order so they chart correctly
df$day_of_week <- factor(df$day_of_week, 
                         levels=c("Sunday", "Monday", "Tuesday", "Wednesday",
                                  "Thursday", "Friday", "Saturday"))
# Create boxplot
boxplot(attend ~ day_of_week, df, xlab="Day", ylab="Attendance", 
        main="Boxplots of Attendance by Day of Week", cex.axis=0.8)
```

Mondays and Wednesdays appear to have room for improvement and may be good candidates for a promotion to increase attendance.

No to look at opponents. There are 17 teams and 81 games, which is fewer than 5 games per team on average, so I don't expect to see much useful information. Perhaps some teams are more popular or are big rivals and we might already expect to see higher attendance on those nights.

```{r opponents}
# Create bar plot showing number of games by opponent
ggplot(df, aes(x=as.factor(opponent))) + geom_bar() + coord_flip() + 
  scale_y_continuous(n.breaks=8) + labs(x="Opponent", y="Number of Games", 
                                        title="Number of Games by Opponent")

# Create violin plots of game attendance by opponent
ggplot(df, aes(x=opponent, y=attend)) + geom_violin() + 
  labs(x="Opponent", y="Attendance", title="Attendance by Opponent") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

There are a handful of opponents that seem to have fairly consistent attendance. Of these, most have only 3-4 games so it's hard to draw conclusions. The only exception might be the Cardinals, who have a fairly consistent attendance over 7 games. Overall, I don't think the opponent should figure into a marketing promotion.

Because our goal is determining which night to hold a promotion, I'm going to skip over weather-related variables 'temp' and 'skies' because we cannot control or predict what the weather will be, apart from seasonal changes, which should be reflected in the 'month' variable.

I don't know what we'll see with day games vs night games.

```{r time of day}
# Boxplot of attendance for day vs night games
boxplot(attend ~ day_night, df, xlab="Time of Day", ylab="Attendance", 
        main="Boxplots of Attendance: Day vs Night", cex.axis=0.8)

# Plot day vs night games by day of week
ggplot(df, aes(x=day_of_week, y=attend, color=day_night)) + geom_point() +
  labs(x="Day of Week", y="Attendance", title="Weekday Attendance by Time") + 
  geom_jitter(width=0.1) + guides(color=guide_legend(title="Time of Day"))
```

There doesn't appear to be much difference in attendance between day and night games and it appears that a decent rule of thumb would be that Day games happen on Sundays, while the others are Night games. Given that we're mostly going to be looking at Mondays and Wednesdays anyway, the time of the game should not be taken into account for the promotion.

Other promotions: caps, shirts, fireworks, and bobbleheads.

```{r other promotions}
# Plot caps vs attendance
ggplot(df, aes(x=cap, y=attend)) + geom_point() + geom_jitter(width=0.1) +
  labs(x="Caps Promotion", y="Attendance", title="Attendance by Caps")

# Plot shirts vs attendance
ggplot(df, aes(x=shirt, y=attend)) + geom_point() + geom_jitter(width=0.1) +
  labs(x="Shirt Promotion", y="Attendance", title="Attendance by Shirts")

# Plot fireworks vs attendance
ggplot(df, aes(x=fireworks, y=attend)) + geom_point() + geom_jitter(width=0.1) +
  labs(x="Fireworks", y="Attendance", title="Attendance by Fireworks")

# Plot bobbleheads vs attendance
ggplot(df,aes(x=bobblehead, y=attend)) + geom_point() + geom_jitter(width=0.1) +
  labs(x="Bobbleheads", y="Attendance", title="Attendance by Bobbleheads")
```

Shirts and Caps promotions are too few in number to draw conclusions, though shirts may correlate slightly with increase attendance, while fireworks don't seem to have much impact. Bobbleheads on the other hand appear to correlate pretty significantly with attendance. What we don't know is whether people are coming out because they love bobblehead promotions or whether bobbleheads are only distributed on nights with high attendance. This is definitely something to look into since we're interested in increasing attendance.

So the key variables we've honed in on are the month of the year ('month'), the day of the week ('day_of_week'), and whether or not bobbleheads were offered ('bobblehead'). Let's put together a regression with these variables.

```{r regression}
# Create linear regression model
relation <- lm(formula = attend ~ month + day_of_week + bobblehead, data = df)
summary(relation)
```

As expected, May and October are expected to have lower attendance, along with the days Monday and Wednesday, but also Thursday, which I hadn't noticed earlier. Bobbleheads meanwhile appear to have a significant impact on attendance.

Let's use this model to predict attendance.

```{r predictions}
# Set up Data Frame of predictions
noBobble <- data.frame(month=c("MAY", "MAY", "MAY", "OCT", "OCT", "OCT"), 
                   day_of_week=c("Monday", "Wednesday", "Thursday"),
                   bobblehead="NO")
# Predict attendance
noBobble$prediction <- predict(relation, noBobble)

# View predictions
noBobble
```

It would appear that Mondays in May (when there are no bobblehead promotions) are expected to have the lowest predicted attendance and should be a good candidate for increasing attendance via promotions. 

If in fact it is the bobbleheads that are bringing people out to the games, that might be a good option for these nights. Let's predict the attendance with a bobblehead promotion on Mondays in May.

``` {r bobbles}
# Create prediction Data Frame
bobble <- data.frame(month="MAY", day_of_week="Monday", bobblehead="YES")

# Predict attendance
bobble$prediction <- predict(relation, bobble)

# View prediction
bobble
```

Adding bobbleheads to the mix accounts for over 10,000 in additional attendance. So, if causitive, this may be the best promotion, assuming the bobbleheads are cost effective. Either way, we should focus on **Mondays in May** for our additional promotional activities.
